---
title: "eda"
author: "Carolina Montes Garcia"
date: "2024-10-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(haven)

```


```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())
```


```{r}
weather_df |> 
  ggplot(aes(x = prcp)) + 
  geom_histogram()
```
  

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) + 
  geom_point(alpha = .75)
```


```{r}
weather_df |>
  group_by(name, month) |>
  summarize(n_obs = n())
```

```{r}
weather_df |>
  pull(month) |> 
  table()
```

```{r}
weather_df |>
  group_by(month) |>
  summarize(
    n_obs = n(),
    n_days = n_distinct(date))
```

```{r}
weather_df |> 
  drop_na(tmax) |> 
  mutate(
    cold = case_when(
      tmax <  5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE      ~ ""
  )) |> 
  filter(name != "Molokai_HI") |> 
  group_by(name, cold) |> 
  summarize(count = n())
```

```{r}
weather_df |> 
  drop_na(tmax) |> 
  mutate(cold = case_when(
    tmax <  5 ~ "cold",
    tmax >= 5 ~ "not_cold",
    TRUE     ~ ""
  )) |> 
  filter(name != "Molokai_HI") |> 
  janitor::tabyl(name, cold)
```

```{r}
weather_df |>
  group_by(month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prec = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax),
    sd_tmax = sd(tmax))
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax),
    median_tmax = median(tmax))
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(across(tmin:prcp, mean))
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax) |> 
  knitr::kable(digits = 1)
```

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax) |> 
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point() 
```

New stuff 'Window Functions'

I can ask for a new variable ranking observations in numerical order
```{r}
weather_df |>
  group_by(name, month) |>
  mutate(temp_ranking = min_rank(tmax))

weather_df |>
  group_by(name) |>
  mutate(temp_rank = min_rank(tmax)) |> 
  filter(temp_rank < 4)

```

Another way of arranging with ranking but without mutate. That bit of code inside filter can serve the same purpose.
```{r}
weather_df |>
  group_by(name) |>
  filter(min_rank(tmax) < 4)|>
  arrange(tmax)
```

The lag function
```{r}
weather_df|>
  mutate(
    lagged_temp = lag(tmax)
  )
```


median bdi score in all 4 visits in the PULSE dataset - for some reason R isn't recognizing the new column 'visit', eventhough I clearly see it in the new dataset for pulse_df...

```{r eval=FALSE}

pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m, 
    names_to = "visit", 
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  )

pulse_df = 
  group_by(visit) %>% 
  summarize(
    mean_bdi = mean(bdi_score, na.rm - TRUE)
  ) %>% 
  knitr: kable(digits = 1)
```

##FAS

```{r}
litter_data = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

pup_data = 
  read_csv("./data/FAS_pups.csv", na = c("NA", ".", "")) |>
  janitor::clean_names()


fas_data = left_join(pup_data, litter_data, by = "litter_number") 

fas_data |> 
  group_by(dose, day_of_tx) |> 
  drop_na(dose) |> 
  summarize(mean_pivot = mean(pd_pivot, na.rm = TRUE)) |> 
  pivot_wider(
    names_from = dose, 
    values_from = mean_pivot) |> 
  knitr::kable(digits = 3)
```

