---
title: "viz_and_eda_@"
author: "Carolina Montes Garcia"
date: "2024-10-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(patchwork)
```

Import the weather data

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Export data -- some folks had trouble accessing.

```{r, eval = FALSE}
write_csv(weather_df, file = "data/weather_df.csv")
```


Making our first plot :-)

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
```


```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```


```{r}
ggp_weather_scatterplot = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

ggp_weather_scatterplot
```

Check why some rows are missing.

```{r}
weather_df |> 
  filter(is.na(tmax))
```


## Fancier scatterplots!

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Where you define aesthetics can matter


```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

use faceting real quick

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

Let's make a somewhat more interesting scatterplot

```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) + 
  geom_point(alpha = .3) + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```


LA plot:

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32
  ) |> 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```


## Small things

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()
```


```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(color = "blue")
```

## Univariate plots

```{r}
weather_df |> 
  ggplot(aes(x = tmin)) + 
  geom_histogram()
```

```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram(position = "dodge")
```

how would i fix this? maybe facet?

```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram() + 
  facet_grid(. ~ name)
```

maybe a density plot?

```{r}
weather_df |> 
  ggplot(aes(x = tmin, fill = name)) + 
  geom_density(alpha = .3)
```

```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmin, fill = name)) + 
  geom_boxplot()
```

violin plots

```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmin, fill = name)) + 
  geom_violin()
```

ridge plot

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = name)) +
  geom_density_ridges()
```


LA univariate plots

```{r}
weather_df |> 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .3)

weather_df |> 
  ggplot(aes(x = name, y = prcp)) + 
  geom_boxplot()

weather_df |> 
  filter(prcp > 10, prcp < 1000) |>  
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .3)
```



## Saving and embedding plots

saving plots

```{r}
ggp_weather = 
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()


ggsave("plots/ggp_weather.pdf", ggp_weather, width = 8, height = 6)
```


embedding plots

```{r, fig.width = 6, fig.asp = .6}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()
```

Make a scatterplot but fancy
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")
```

Scaling

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_color_hue(h = c(100, 300))


ggp_temp_plot = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )

ggp_temp_plot
```

Legend at the bottom instead of the right
```{r}
ggp_temp_plot + 
  theme(legend.position = "bottom")
```

Modify background theme
```{r}
ggp_temp_plot + 
  theme_bw() + 
  theme(legend.position = "bottom")

ggp_temp_plot + 
  theme_minimal() + 
  theme(legend.position = "bottom")

ggp_temp_plot + 
  theme_classic() + 
  theme(legend.position = "bottom")
```

Learning check
SE are turned off 
```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) + 
  geom_point(aes(size = prcp), alpha = .75) + 
  labs(
    title = "Temperature plot",
    x = "Date",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

Extra bonus stuff in 'ggplot'

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park_df) 
```

##Multiple panels

```{r}
tmax_tmin_p = 
  weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df |> 
  filter(prcp > 0) |> 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) + 
  theme(legend.position = "none")

tmax_date_p = 
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme(legend.position = "bottom")

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```

